---
title: Webappendix for the article entitled Insights for quantifying the long-term
  benefit of immunotherapy using quantile regression
author: "Bassirou Mboup, Christophe Le Tourneau,  Aurélien Latouche"
date: "04/05/2020"
output:
  pdf_document: default
  bibliography: bibli.bib 
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(pandocfilters)
library(car)
library(survival)
library(quantreg)
library(survminer)
library(boot)
library(ggplot2)

```


This document makes an easier access to the supplementary material of the article entitled **Insights for quantifying the long-term benefit of immunotherapy using quantile regression**.

## 1) Importing the reconstructed data set

We use the algorithm of Guyot. al 2012 to reconstruct individual-level time-to-event data based on the published Kaplan–Meier curves of the randomized controlled trial (Rittmeyer et al. 2017). 

The R code of the algorithm is available at https://www.mskcc.org/sites/default/files/node/137932/documents/2017-04-20-14-31-36/dataexample.zip 

After reconstruction, we get in this dataset the following variables.

- time : vector of observed failure times e.g (death or censored).

- event: vector of indicator of status (1 for death and 0 for censoring).

- tmt.arm.number: vector of treatment indicator (binary with 1 for immunotherapy).

- treatment.type: the type of treatment (char) immonotherapy or chemotherapy

```{r echo=FALSE}
load("data_ICI_Rittmeyer.RData")
head(data_ICI_Rittmeyer)
data_ICI_Rittmeyer$tmt.arm.number <- ifelse(data_ICI_Rittmeyer$treatment.type=="Atezolizumab",1,0)
```
```{r}
## Kaplan Meier curves 
fit_KM <- survfit(Surv(time,event)~tmt.arm.number,data=data_ICI_Rittmeyer)

res <- ggsurvplot(fit_KM,data=data_ICI_Rittmeyer,
           risk.table=TRUE,
           conf.int=FALSE,
           xlim=c(0.4,27),
           palette =c("red","blue"),
           xlab="Months",
           risk.table.y.text.col=T,
            break.time.by=3,
          ggtheme = theme_grey() ,
           legend.title="Treatment",
           legend.labs=c("Chemo","Immuno")
            )
res$table <- res$table + theme(axis.line = element_blank())
res$plot <- res$plot+geom_hline(yintercept=0.28,lty=2)
print(res)

```



## 2) Application of quantile regression for survival data
In the following R output,
test of the coefficients compares this coefficients to $0$ for a given quantile $\tau$. For the coefficient $\beta_1(\tau)$, this test rejects or not the equality of the two treatment groups for a given quantile $\tau$. This test is based of the Wald test and the variance of the coefficients is obtained using resampling bootstraps method.

Thus, we do not reject the hypothesis of equality of the two groups at 0.1 quantile  (P.value=0.9470) illustrated by the overlap of the curves at this quantile.  

We reject this hypothesis at 0.6 quantile (P.value $<10^{-4}$).


```{r warning=FALSE}
set.seed(123456)
x <- c(0.1, 0.2, 0.3, 0.4,0.5,0.6)
Rq <- crq(Surv(time,event)~tmt.arm.number,data=data_ICI_Rittmeyer,method="Pen") 
result <- summary(Rq,taus=x)

result

# Quantile function 
tau <- Rq$sol["tau",][1:160]
q<- Rq$sol["Qhat",][1:160]
plot(tau,q,type="l",xlab = expression(tau),ylab = expression(QT(tau)), 
     main="Quantile function",axes=FALSE)
axis(1,at=seq(from=0,to=0.7,by=0.1),labels=seq(from=0,to=0.7,by=0.1),las=1)
axis(2,at=seq(from=0,to=20,by=5),labels=seq(from=0,to=20,by=5),las=2)

```




## 3) Resampling for confidence interval of  coefficients  

```{r}

############ data for bootstrapping with 1000 replications ################
load("Mat.RData") 
load("Moy.RData")
load("Mat1.RData")
load("Moy1.RData")

b <- matrix(data=NA,ncol=6,nrow = 1000)
for(i in 1:6){
b[,i] <-Mat1[,i][order(Mat1[,i])]
}

inf0 <- NA
for(i in 1:6){
inf0[i] <- b[,i][25]
}
sup0 <- NA
for(i in 1:6){
  sup0[i] <- b[,i][975]
}

plot(x,Moy1,type="l",ylim=c(0,16),ylab="coefficients",xlab ="Quantile",main="Intercept",axes = FALSE)
axis(1,at=seq(from=0.1,to=0.6,by=0.1),labels=seq(from=0.1,to=0.6,by=0.1),las=1)
axis(2,at=seq(from=0,to=16,by=2),labels=seq(from=0,to=16,by=2),las=2)
lines(x,sup0,col="blue",lty=2)
lines(x,inf0,col="blue",lty=2)

```




```{r}
## Variation of the treatment effect as function of each quantile available and confidence 
# intervals
b <- matrix(data=NA,ncol=6,nrow = 1000)
for(i in 1:6){
b[,i] <-Mat[,i][order(Mat[,i])]
}

inf <- NA
for(i in 1:6){
inf[i] <- b[,i][25]
}
sup <- NA
for(i in 1:6){
  sup[i] <- b[,i][975]
}


plot(x,Moy,type="l",ylim=c(-2,12),ylab="coefficients",xlab ="Quantile",main="Treatment effect",axes = FALSE)
axis(1,at=seq(from=0.1,to=0.6,by=0.1),labels=seq(from=0.1,to=0.6,by=0.1),las=1)
axis(2,at=seq(from=-2,to=12,by=2),labels=seq(from=-2,to=12,by=2),las=2)
lines(x,sup,col="blue",lty=2)
lines(x,inf,col="blue",lty=2)
```


## 4) Benefit of treatment for each quantile

```{r}
datafr <- rbind.data.frame(Moy,inf,sup)
names(datafr) <- c( "0.1", "0.2", "0.3", "0.4", "0.5","0.6")
p <- ggplot(stack(datafr), aes(x=factor(ind,levels=names(datafr)), y=values))+geom_boxplot()
p+labs(x="Quantile",y="Benefit in months")
```




## 5) Visual check  of the linearity assumption 

```{r}
x <- c(0.1,0.2,0.3,0.4,0.5,0.6)
Rq <- crq(Surv(time,event)~tmt.arm.number,data=data_ICI_Rittmeyer,method="Pen")
tau <- Rq$sol["tau",][1:160]
obs <- unname(quantile(data_ICI_Rittmeyer$time,tau))
obss <- unname(quantile(data_ICI_Rittmeyer$time,x))
q<- Rq$sol["Qhat",][1:160]
q1 <- Rq$sol["Qhat",][c(24,46,69,91,114,136)]
residu <- obs-q
residus <- obss-q1
scatterplot(q,log(abs(residu)),xlab="fitted quantile value",ylab="Martingal residuals")
points(q1,log(abs(residus)),col='red')
for(i in 1:6){
  abline(v=q1[i],col="red")
}


```


## 6) Testing equality of two groups for  a given quantile using survival Kaplan Meier function 


For a given quantile, we can rely on 2-samples test derived for the median survival as detailed below.
Testing for equality of median was derived in (Tang et al.,Chen et al.)
Once the desired quantile is identified, the methodology can be adapted. 
These tests,  designed for detecting the difference of the median survival times,  can be readily  extended  to compare survival quantiles. 

Therefore, let's assume that 
\begin{align*}
\widehat{F}_1^{-1}(q)&=\inf\{t:\widehat{F}_1(t)=1-\widehat{S}_1(t)\geq q\}\\
\widehat{F}_2^{-1}(q)&=\inf\{t:\widehat{F}_2(t)=1-\widehat{S}_2(t)\geq q\} , \forall q\in [0,1]
\end{align*}
where $\widehat{S}_1$ and $\widehat{S}_2$ are respectively the estimated of the survival functions $S_1$ for non treated groups and $S_2$ for treated groups using the Kaplan Meier method, $widehat{F}_1$ and $\widehat{F}_2$ are right continuous, piece-wise constant estimators of $F_1$ and $F_2$ respectively.

Testing the equality of the quantile between the two groups is equivalent to testing the null hypothesis
\begin{align*}
H_0: F_1^{-1}(q)&=F_2^{-1}(q)\\
     F_2\{F_1^{-1}(q)\}&=q
\end{align*}
As pointed out by Kosorok et al., $\sqrt{n}(F_2\{F_1^{-1}(q)\}-q)$ is asymptotically a zero-mean Gaussian process with variance $\sigma^2$. 

We estimated the variance  $\hat{\sigma}^2$ using re-sampling bootstrap method. The following statistic test 
\begin{align*}
\frac{(\widehat{F}_2\{\widehat{F}_1^{-1}(q)\}-q)^2}{\sigma^2}
\end{align*}
follows a $\chi^2$-distribution with 1 degrees of freedom.

We applied this test in our data set at the quantile level $0.6$ highly significant  with a  p-value $< 10^{-4}$ of 
%$1.892087e^{-05}$,
which indicates a significant difference at the quantile level $0.6$ survival time between the two treatment groups.




## R code 

```{r}

quantileTest <- function(time,event,treat,q,B=1000,seed=1234){
  set.seed(seed)
  Mesdon <- cbind.data.frame(time=time,event=event,treat=treat)
  fit1 <- survfit(Surv(Mesdon$time[Mesdon$treat==0],Mesdon$event[Mesdon$treat==0])~1,conf.type ="none")
  fit2 <- survfit(Surv(Mesdon$time[Mesdon$treat==1],Mesdon$event[Mesdon$treat==1])~1,conf.type = "none")
  F1.inv <- unname(quantile(fit1, prob = q)) 
  F2.inv <- unname(quantile(fit2, prob = q))
  
  # Calculate F2(F1.inv(p))
  Qp <- function(t1,c1, t2, c2) {
    fit1 <- survfit(Surv(t1, c1)~1, conf.type = "none")
    fit2 <- survfit(Surv(t2, c2)~1, conf.type = "none")
    F1.inv <- unname(quantile(fit1, prob=q)) 
    if (is.na(F1.inv)) {
      warning(paste0("Error"))
      F1.inv <- max(t1)
    }
    F2  <- stepfun(fit2$time, c(0, 1-fit2$surv)) #CDF of F2
    out <- F2(F1.inv) #F2(F1.inv(p))
    return(out)
  }

  Q <- Qp(Mesdon$time[Mesdon$treat==0],Mesdon$event[Mesdon$treat==0],
          Mesdon$time[Mesdon$treat==1],Mesdon$event[Mesdon$treat==1])
  
  # Bootstrap
  b.est <- numeric(B)
  for (i in 1:B) {
    boot1    <- sample(1:length(Mesdon$time[Mesdon$treat==0]),replace =TRUE)
    t1.boot  <- Mesdon$time[Mesdon$treat==0][boot1]
    c1.boot  <- Mesdon$event[Mesdon$treat==0][boot1]
    boot2    <- sample(1:length(Mesdon$time[Mesdon$treat==1]),replace = TRUE)
    t2.boot  <- Mesdon$time[Mesdon$treat==1][boot2]
    c2.boot  <- Mesdon$event[Mesdon$treat==1][boot2]
    b.est[i] <- Qp(t1.boot,c1.boot,t2.boot,c2.boot)
  }
  
  se   <- sd(b.est)
  Z<- (Q-q)^2/se^2
  pval <- 1-pchisq(Z,1)
  return(pval)
}


## Application of the test with our data at 0.6 quantile
quantileTest(time=data_ICI_Rittmeyer$time,event=data_ICI_Rittmeyer$event,
             treat = data_ICI_Rittmeyer$tmt.arm.number,q=0.6)

```


