---
title: Webappendix for the article entitled Insights for quantifying the long-term
  benefit of immunotherapy using quantile regression
author: "Bassirou Mboup, Christophe Le Tourneau,  Aurélien Latouche"
date: "07/05/2020"
output:
  pdf_document: default
  bibliography: bibli.bib 
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(pandocfilters)
library(car)
library(survival)
library(quantreg)
library(survminer)
library(boot)
library(ggplot2)

```


This document makes an easier access to the supplementary material of the article entitled **Insights for quantifying the long-term benefit of immunotherapy using quantile regression**.

## 1) Importing the reconstructed data set

We use the algorithm of Guyot. al 2012 to reconstruct individual-level time-to-event data based on the published Kaplan–Meier curves of the randomized controlled trial (Rittmeyer et al. 2017). 

The R code of the algorithm is available at https://www.mskcc.org/sites/default/files/node/137932/documents/2017-04-20-14-31-36/dataexample.zip 

After reconstruction, we get in this dataset the following variables.

- time : vector of observed failure times e.g (death or censored).

- event: vector of indicator of status (1 for death and 0 for censoring).

- tmt.arm.number: vector of treatment indicator (binary with 1 for immunotherapy).

- treatment.type: the type of treatment (char) immonotherapy or chemotherapy

```{r echo=FALSE}
load("data_ICI_Rittmeyer.RData")
head(data_ICI_Rittmeyer)
data_ICI_Rittmeyer$tmt.arm.number <- ifelse(data_ICI_Rittmeyer$treatment.type=="Atezolizumab",1,0)

```



```{r}
## Kaplan Meier curves 
fit_KM <- survfit(Surv(time,event)~tmt.arm.number,data=data_ICI_Rittmeyer)

res <- ggsurvplot(fit_KM,data=data_ICI_Rittmeyer,
           risk.table=TRUE,
           conf.int=FALSE,
           xlim=c(0.4,27),
           palette =c("red","blue"),
           xlab="Months",
           risk.table.y.text.col=T,
            break.time.by=3,
          ggtheme = theme_grey() ,
           legend.title="Treatment",
           legend.labs=c("Chemo","Immuno")
            )
res$table <- res$table + theme(axis.line = element_blank())
res$plot <- res$plot+geom_hline(yintercept=0.28,lty=2)
print(res)

```



## 2) Application of quantile regression for survival data
In the following R output,
test of the coefficients compares this coefficients to $0$ for a given quantile $\tau$. For the coefficient $\beta_1(\tau)$, this test rejects or not the equality of the two treatment groups for a given quantile $\tau$. This test is based of the Wald test and the variance of the coefficients is obtained using resampling bootstraps method.

Thus, we do not reject the hypothesis of equality of the two groups at 0.1 quantile  (P.value=0.9470) illustrated by the overlap of the curves at this quantile.  

We reject this hypothesis at 0.6 quantile (P.value $<10^{-4}$).


```{r warning=FALSE}
set.seed(123456)
x <- c(0.1, 0.2, 0.3, 0.4,0.5,0.6)
Rq <- crq(Surv(time,event)~tmt.arm.number,data=data_ICI_Rittmeyer,method="Pen") 
result <- summary(Rq,taus=x)

result

# Quantile function 
tau <- Rq$sol["tau",][1:160]
q<- Rq$sol["Qhat",][1:160]
plot(tau,q,type="l",xlab = expression(tau),ylab = expression(QT(tau)), 
     main="Quantile function",axes=FALSE)
axis(1,at=seq(from=0,to=0.7,by=0.1),labels=seq(from=0,to=0.7,by=0.1),las=1)
axis(2,at=seq(from=0,to=20,by=5),labels=seq(from=0,to=20,by=5),las=2)

```




## 3) Resampling for confidence interval of  coefficients  

```{r}

############ data for bootstrapping with 1000 replications ################
load("Mat.RData") 
load("Moy.RData")
load("Mat1.RData")
load("Moy1.RData")

b <- matrix(data=NA,ncol=6,nrow = 1000)
for(i in 1:6){
b[,i] <-Mat1[,i][order(Mat1[,i])]
}

inf0 <- NA
for(i in 1:6){
inf0[i] <- b[,i][25]
}
sup0 <- NA
for(i in 1:6){
  sup0[i] <- b[,i][975]
}

plot(x,Moy1,type="l",ylim=c(0,16),ylab="coefficients",xlab ="Quantile",main="Intercept",axes = FALSE)
axis(1,at=seq(from=0.1,to=0.6,by=0.1),labels=seq(from=0.1,to=0.6,by=0.1),las=1)
axis(2,at=seq(from=0,to=16,by=2),labels=seq(from=0,to=16,by=2),las=2)
lines(x,sup0,col="blue",lty=2)
lines(x,inf0,col="blue",lty=2)

```




```{r}
## Variation of the treatment effect as function of each quantile available and confidence 
# intervals
b <- matrix(data=NA,ncol=6,nrow = 1000)
for(i in 1:6){
b[,i] <-Mat[,i][order(Mat[,i])]
}

inf <- NA
for(i in 1:6){
inf[i] <- b[,i][25]
}
sup <- NA
for(i in 1:6){
  sup[i] <- b[,i][975]
}


plot(x,Moy,type="l",ylim=c(-2,12),ylab="coefficients",xlab ="Quantile",main="Treatment effect",axes = FALSE)
axis(1,at=seq(from=0.1,to=0.6,by=0.1),labels=seq(from=0.1,to=0.6,by=0.1),las=1)
axis(2,at=seq(from=-2,to=12,by=2),labels=seq(from=-2,to=12,by=2),las=2)
lines(x,sup,col="blue",lty=2)
lines(x,inf,col="blue",lty=2)
```


## 4) Benefit of treatment for each quantile

```{r}
datafr <- rbind.data.frame(Moy,inf,sup)
names(datafr) <- c( "0.1", "0.2", "0.3", "0.4", "0.5","0.6")
p <- ggplot(stack(datafr), aes(x=factor(ind,levels=names(datafr)), y=values))+geom_boxplot()
p+labs(x="Quantile",y="Benefit in months")
```




## 5) Visual check  of the linearity assumption 

```{r}
x <- c(0.1,0.2,0.3,0.4,0.5,0.6)
Rq <- crq(Surv(time,event)~tmt.arm.number,data=data_ICI_Rittmeyer,method="Pen")
tau <- Rq$sol["tau",][1:160]
obs <- unname(quantile(data_ICI_Rittmeyer$time,tau))
obss <- unname(quantile(data_ICI_Rittmeyer$time,x))
q<- Rq$sol["Qhat",][1:160]
q1 <- Rq$sol["Qhat",][c(24,46,69,91,114,136)]
residu <- obs-q
residus <- obss-q1
scatterplot(q,log(abs(residu)),xlab="fitted quantile value",ylab="Martingal residuals")
points(q1,log(abs(residus)),col='red')
for(i in 1:6){
  abline(v=q1[i],col="red")
}


```




##


```{r}

load("ImmunoRit.RData")
library(knitr)
library(riskRegression)

db <- ImmunoRit
cox1<-coxph(Surv(time,event)~Treat, data=db,x=TRUE)

db2<-mutate(db,Treat=0)

pr<-predictCox(cox1,times=db2$time,newdata=db2,type="cumhazard")
cumhaz<-pr$cumhazard[1,]



beta1 <- log(0.59)
lamda <- cumhaz
beta2 <- log(0.74)
beta3 <- log(0.36)

genereTime <- function(x,lamda,trt,uni){
  return(-log(uni)/(lamda*exp(beta2*trt+beta1*x+beta3*x*trt)))
}

ttt <- ImmunoRit$Treat
n=850

PDL1 <- rep(NA,n)
for(i in 1:n) {
  x <- -1
  while(x<0){
    uni <- runif(1,0,1)
    x <- uniroot((function(x)(genereTime(x,lamda[i],ttt[i],uni)-ImmunoRit$time[i])),lower=-10,upper=10,extendInt = "yes")$root 
    
  }
  PDL1[i] <- x
}

summary(PDL1)
ImmunoRit$PDL1 <- PDL1

cox<-coxph(Surv(time,event)~Treat*PDL1,data=db,x=TRUE)
summary(cox)

Rq_1 <- crq(Surv(time,event)~Treat*PDL1,data=ImmunoRit,method="PengHuang")
result <- summary(Rq_1,tau=c(0.1,0.20,0.30,0.40,0.50,0.6))
kable(result[[6]][2]$coefficients)

```



